The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-08-14 talk/vmil2012/paper.tex
Remove most of \texttt annotations

The operations executed by an interpreter are recorded by the tracing JIT in
case they are frequently executed (this process is described in more detail in
Section \ref{sec:Resume Data}). During the recording phase \texttt{guards} are
inserted into the recorded trace at all
points where the control flow could diverge. As can be seen in
Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
operations before and for about 15\% to 20\% of the operations after
optimizing the traces generated for the different benchmarks used in this paper.
This makes guards one of the most common types of operations. As this paper will show, many of these guards
fail rarely or not all during execution.

@@

As explained in previous sections, when a specific guard has failed often enough
a new trace, referred to as a \emph{bridge}, starting from this guard is recorded and
compiled.
Since the goal of compiling bridges is to improve execution speed on the
diverged path (failing guard) they should not introduce additional overhead.
In particular the failure of the guard should not lead
to leaving the compiled code prior to execution the code of the bridge.

@@

Once the bridge has been compiled the corresponding guard is
patched to redirect control flow to the bridge in case the check fails. In
the future, if the guard fails again it jumps to the code compiled for the bridge
instead of bailing out. Once the guard has been compiled and attached to the
loop the guard becomes just a point where control-flow can split. The loop
after the guard and the bridge are just conditional paths.
Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
Guard \#1 jumps to the trampoline, loads the \texttt{backend map} and
then calls the bailout handler, whereas Guard \#2 has already been patched
and directly jumps to the corresponding bridge. The bridge also contains two
guards that work based on the same principles.
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/loop_bridge.pdf}
\caption{Trace control flow in case of guard failures with and without bridges}
\label{fig:trampoline}
\end{figure}
% section Guards in the Backend (end)

@@

The overhead that is incurred by the JIT to manage the \texttt{resume data},
the \texttt{backend map} as well as the generated machine code is
shown in Figure~\ref{fig:backend_data}. It shows the total memory consumption
of the code and of the data generated by the machine code backend and an
approximation of the size of the \texttt{resume data} structures for the
different benchmarks mentioned above. The machine code taken into account is
composed of the compiled operations, the trampolines generated for the guards
and a set of support functions that are generated when the JIT starts and which
are shared by all compiled traces. The size of the \texttt{backend map}
is the size of the compressed mapping from registers and stack to
IR-level variables and finally the size of the \texttt{resume data} is an
approximation of the size of the compressed high-level resume data as described
in Section~\ref{sec:Resume Data}.\footnote{
The size of the resume data is not measured at runtime, but reconstructed from
log files.}

For the different benchmarks the \texttt{backend map} has a size of
about 15\% to 20\% of the amount of memory compared to the size of the
generated machine code. On the other hand the generated machine code has only a
size ranging from 20.5\% to 37.98\% of the size of the resume data and the backend map
combined and being compressed as described before.

Tracing JIT compilers only compile the subset of the code executed in a program
that occurs in a hot loop, for this reason the amount of generated machine
code will be smaller than in other juts-in-time compilation approaches.  This
creates a larger discrepancy between the size of the \texttt{resume data} when
compared to the size of the generated machine code and illustrates why it is important to compress the \texttt{resume data} information.

@@

Why the efficient storing of the \texttt{resume data} is a central concern in the design
of guards is illustrated by Figure~\ref{fig:resume_data_sizes}. This figure shows
the size of the compressed \texttt{resume data}, the approximated size of
storing the \texttt{resume data} without compression and
an approximation of the best possible compression of the resume data by
compressing the data using the
\texttt{xz} compression tool, which is a ``general-purpose data compression
software with high compression ratio''.\footnote{\url{http://tukaani.org/xz/}}

>>>>

The operations executed by an interpreter are recorded by the tracing JIT in
case they are frequently executed (this process is described in more detail in
Section \ref{sec:Resume Data}). During the recording phase guards are
inserted into the recorded trace at all
points where the control flow could diverge. As can be seen in
Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
operations before and for about 15\% to 20\% of the operations after
optimizing the traces generated for the different benchmarks used in this paper.
This makes guards one of the most common types of operations. As this paper will show, many of these guards
fail rarely or not all during execution.

@@

As explained in previous sections, when a specific guard has failed often enough
a bridge starting from this guard is recorded and compiled.
Since the goal of compiling bridges is to improve execution speed on the
diverged path (failing guard) they should not introduce additional overhead.
In particular the failure of the guard should not lead
to leaving the compiled code prior to execution the code of the bridge.

@@

Once the bridge has been compiled the corresponding guard is
patched to redirect control flow to the bridge in case the check fails. In
the future, if the guard fails again it jumps to the code compiled for the bridge
instead of bailing out. Once the guard has been compiled and attached to the
loop the guard becomes just a point where control-flow can split. The loop
after the guard and the bridge are just conditional paths.
Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
Guard \#1 jumps to the trampoline, loads the backend map and
then calls the bailout handler, whereas Guard \#2 has already been patched
and directly jumps to the corresponding bridge. The bridge also contains two
guards that work based on the same principles.
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/loop_bridge.pdf}
\caption{Trace control flow in case of guard failures with and without bridges}
\label{fig:trampoline}
\end{figure}
% section Guards in the Backend (end)

@@

The overhead that is incurred by the JIT to manage the resume data,
the backend map as well as the generated machine code is
shown in Figure~\ref{fig:backend_data}. It shows the total memory consumption
of the code and of the data generated by the machine code backend and an
approximation of the size of the resume data structures for the
different benchmarks mentioned above. The machine code taken into account is
composed of the compiled operations, the trampolines generated for the guards
and a set of support functions that are generated when the JIT starts and which
are shared by all compiled traces. The size of the backend map
is the size of the compressed mapping from registers and stack to
IR-level variables and finally the size of the resume data is an
approximation of the size of the compressed high-level resume data as described
in Section~\ref{sec:Resume Data}.\footnote{
The size of the resume data is not measured at runtime, but reconstructed from
log files.}

For the different benchmarks the backend map has a size of
about 15\% to 20\% of the amount of memory compared to the size of the
generated machine code. On the other hand the generated machine code has only a
size ranging from 20.5\% to 37.98\% of the size of the resume data and the backend map
combined and being compressed as described before.

Tracing JIT compilers only compile the subset of the code executed in a program
that occurs in a hot loop, for this reason the amount of generated machine
code will be smaller than in other juts-in-time compilation approaches.  This
creates a larger discrepancy between the size of the resume data when
compared to the size of the generated machine code and illustrates why it is important to compress the resume data information.

@@

Why the efficient storing of the resume data is a central concern in the design
of guards is illustrated by Figure~\ref{fig:resume_data_sizes}. This figure shows
the size of the compressed resume data, the approximated size of
storing the resume data without compression and
an approximation of the best possible compression of the resume data by
compressing the data using the
\emph{xz} compression tool, which is a ``general-purpose data compression
software with high compression ratio''.\footnote{\url{http://tukaani.org/xz/}}
